Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex) with SVA adjustments

### 2016-06-11 code/Rmd/Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd
Added markdown to process Mayo RNASeq based on PMI guided SVA

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_SVA_Reprocessed.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)')
```

@@

thisFileName <- 'Mayo_RNASeq_SVA_Reprocessed.Rmd'

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design

## Calculate the number of surrogate variables
# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]

# Permute and calculate the permuted singular values
PERM.SING.VALS <- foreach (i = 1:100,
         .combine = rbind,
         .packages = c('limma','CovariateAnalysis'),
         .export = c('EXPR','MODEL1')) %dopar% {

           # Permute expression counts
           PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))

           # Estimate voom weights
           VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)

           # Fit linear model using new weights and new design
           VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)

           # Residuals after normalisation
           RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)

           PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
           PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
        }
num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
NUM.SV = min(num.sv1, num.sv2)

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, method = "two-step", B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

@@

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID)))
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA adjustments)', parentId = parentId)
CODE <- synStore(CODE)

@@

@@

# Store differential expression results
write.table(DE, file = 'MAYOdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYOdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
stopCluster(cl)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r COV_OBJ$properties$id`      |
|  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
|  Residual Expression      |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`     |

>>>>

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA based PMI adjustments)')
```

@@

thisFileName <- 'Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd'

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
# Get genes that are predictors of PMI and use them as control probes
CONTROL.PROBES_ID = 'syn6145639'
ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F)
cntrl = rep(0.5, dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
names(cntrl) = sapply(rownames(VOOM.ADJUSTED.GENE_EXPRESSION$E), function(x){unlist(str_split(x,'\\.'))[1]})
cntrl[intersect(unlist(CONTROL.PROBES), names(cntrl))] = 1

# Get actual and null model
MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design

# ## Calculate the number of surrogate variables
# # Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
# 
# # Permute and calculate the permuted singular values
# PERM.SING.VALS <- foreach (i = 1:100,
#          .combine = rbind,
#          .packages = c('limma','CovariateAnalysis'),
#          .export = c('EXPR','MODEL1')) %dopar% {
#            
#            # Permute expression counts
#            PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
#            
#            # Estimate voom weights
#            VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
#            
#            # Fit linear model using new weights and new design
#            VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
#            
#            # Residuals after normalisation
#            RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
#            
#            PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
#            PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
#         }
# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
NUM.SV = num.sv2; #min(num.sv1, num.sv2)

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, 
               MODEL1, MODEL0, 
               n.sv = NUM.SV, 
               controls = cntrl,
               method = "supervised", 
               B = 30)

SURR.VAR = SURR.VAR$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

@@

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID)))
kable(tmp)

DE = DE %>%
  left_join(backgroundGenes) %>% left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, Comparison, BrainRegion, Name, 
                logFC, AveExpr, t, P.Value, adj.P.Val, B)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Reprocessed Counts (with SVA based PMI adjustments)', parentId = parentId)
CODE <- synStore(CODE)

@@

# Store surrogate variables
rownames(SURR.VAR) = rownames(COVARIATES)
SURR.VAR = rownameToFirstColumn(SURR.VAR, 'ID')
write.table(SURR.VAR, file = 'MAYOsurrogatevariable.tsv', sep = '\t', row.names=F, quote=F)
SURR_OBJ = File('MAYOsurrogatevariable.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
SURR_OBJ = synStore(SURR_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)

@@

# Store differential expression results
write.table(DE, file = 'MAYOdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYOdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
stopCluster(cl)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r COV_OBJ$properties$id`      |
|  Surrogate Variables      |  `r SURR_OBJ$properties$id`     |
|  Raw counts (logCPM)      |  `r RAWCPM_OBJ$properties$id`   |
|  Residual Expression      |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`     |
